The YUIMA Project: a Computational Framework for Simulation and Inference of Stochastic Differential Equations

### 2013-02-07 pkg/yuimadocs/inst/doc/JSS/article-new.Rnw
update
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/yuima@221 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
up
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/yuima@223 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@

From the above, it is possible to see that the jump coefficient is void and the Hurst parameter is set to 0.5, \textcolor{blue}{because this is a model where the driving process is the standard Brownian motion, i.e. a fractional Brownian motion if Hurst index $H=\frac12$.}
Now, with \code{mod1} in hands, it is very easy to simulate a trajectory \textcolor{blue}{Euler-Maruyama scheme} of the process as follows
<<echo=TRUE, print=FALSE,fig=TRUE,width=9,height=4,results=hide>>=
set.seed(123)
X <- simulate(mod1)
plot(X)
@

@@

\subsection{Multidimensional Processes}\label{sec:multi}
\textcolor{blue}{Next is an example of a system of two stochastic differential equations for the couple $(X_{t,1}, X_{t,2})$ driven by three independent Brownian motions $(W_{t,1}, W_{t,2}, W_{t,3})$}
$$
\begin{aligned}
\de X_{1,t} &= -3 X_{1,t} \de t + \de W_{1,t} + X_{2,t} \de W_{3,t}\\
\de X_{2,t} &= -(X_{1,t} + 2 X_{2,t}) \de t + X_{1,t} \de W_{1,t} + 3 \de W_{2,t}
\end{aligned}
$$
but this has to be organized into matrix form \textcolor{red}{with a vector of drift expressions and a diffusion matrix}
$$
\left(\begin{array}{c}\de X_{1,t} \\\de X_{2,t}\end{array}\right)=
\left(\begin{array}{c} -3 X_{1,t} \\  -X_{1,t} - 2X_{2,t}\end{array}\right)\de t +
\left[\begin{array}{ccc}1 & 0 & X_{2,t} \\X_{1,t} & 3 & 0\end{array}\right]
\left(\begin{array}{c}\de W_{t,1} \\ \de W_{t,2} \\ \de W_{t,3}\end{array}\right)
$$
\textcolor{blue}{For this system it now necessary to instruct \pkg{yuima} about the state variable on bot the left-hand side of the equation and the right-hand side of the equation, i.e. the \code{solve.variable}.}
<<echo=TRUE, print=FALSE,results=hide>>=
sol <- c("x1","x2") # variable for numerical solution
a <- c("-3*x1","-x1-2*x2")   # drift vector 
b <- matrix(c("1","x1","0","3","x2","0"),2,3)  #  diffusion matrix
mod3 <- setModel(drift = a, diffusion = b, solve.variable = sol)
@
\textcolor{blue}{Looking at the structure of the \code{noise.number} slot in \code{mod3}, one can see that this is now set to 3 which is taken as the number of columns of the diffusion matrix.}
Again, this model can be easily simulated
<<echo=TRUE, print=FALSE,fig=TRUE,width=9,height=4,results=hide>>=
set.seed(123)
X <- simulate(mod3)
plot(X, plot.type="single",lty=1:2)
@

@@

\section{Fractional Gaussian Noise}\label{sec:fgn}
The \pkg{yuima} allows for the description of stochastic differential equations driven by fractional Brownian motion of the following type
$$\de X_t =  a(X_t) \de t + b(X_t) \de W_t^H
$$
where $W^H=\left( W^H_t, t\geq 0 \right)$ is a normalized fractional Brownian motion (fBM), {\it i.e.} the zero mean Gaussian processes with covariance function
$$E( W_s^H W_t^H )= \frac{1}{2} \left( |s|^{2H}+|t|^{2H}- |t-s|^{2H}\right)$$ with Hurst exponent $H\in(0,1)$.  The fractional Brownian motion process is  neither Markovian nor a semimartingale for $H\neq\frac{1}{2}$ but
remains Gaussian.  For $H > \frac12$, the solution $X_t$ above  presents the long-range dependance property that makes it useful for different applications in biology, physics, ethernet traffic or finance.
In order to specify a stochastic differential equation driven by fractional Gaussian noise it is necessary to specify the value of the Hurst parameter. For example, if we want to specify the following model
$$\de Y_t =  3 Y_t \de t + \de W_t^H$$
we proceed as follows
<<echo=TRUE, print=FALSE,fig=TRUE,width=9,height=6,results=hide>>=
mod4A <- setModel(drift="3*y", diffusion=1, hurst=0.3, solve.var="y")
mod4B <- setModel(drift="3*y", diffusion=1, hurst=0.7, solve.var="y")
set.seed(123)
X1 <- simulate(mod4A,sampling=setSampling(n=1000))
X2 <- simulate(mod4B,sampling=setSampling(n=1000))
par(mfrow=c(2,1))
par(mar=c(2,3,1,1))
plot(X1,main="H=0.3")
plot(X2,main="H=0.7")
@

@@

\section{Asymptotic Expansion}\label{sec4}
{\colord For numerical computation of the expectation of a functional, 
the Monte-Carlo methods universally gives a solution although they are time consuming and involve stochastic error 
depending on the number of repetitions. 
An alternative to the Monte-Carlo method is the method of asymptotic expansion, that can often give a solution with comparable accuracy, 
besides, it is much superior in the computation time because of the approximation by an analytic formula. 
 }

The \pkg{yuima} package {\colord provides numerical approximation to the expectation of 
a functional of a diffusion process in terms of the asymptotic expansion scheme. 
Let us consider a family of}
 $d$-dimensional diffusion processes {\colord $X=(X^{(\ep)}_t)_{t\in[0,T]}$ ($\ep\in(0,1]$) specified by the stochastic differential equasions}
$$\de X_t^{(\ve)} = a(X_t^{(\ve)},\ve)\de t + b(X_t^{(\ve)},\ve)\de W_t, \qquad \ve \in(0,1]$$
{\colord where}  $W_t$ \textcolor{red}{is an} $r$-dimensional Wiener process, i.e., $W_t=(W_{1,t}, \ldots, W_{r,t})$.
The functional is expressed in the following abstract form
\begin{equation}
\label{yuima:func1}
F^{(\ve)}  = \sum_{\alpha=0}^r \int_0^T f_\alpha(X_t^{(\ve)},\de)\de W_t^\alpha + F(X_t^{(\ve)},\ve), \qquad W_t^0=t.
\end{equation} 
A typical example of application is the case of Asian option pricing. For example, in the Black \& Scholes model
\begin{equation}
\label{eqyuima:gbm1}
\de X_t^{(\ve)} = \mu X_t^{(\ve)}\de t + \ve X_t^{(\ve)} \de W_t
\end{equation}
the price of the option is of the form
$$ \dE\left\{ \max\left( \frac{1}{T}\int_0^T X_t^{(\ve)} \de t - K,0\right)\right\}.$$
Thus the functional of interest is
$$F^{(\ve)} = \frac{1}{T}\int_0^T X_t^{(\ve)} \de t, \qquad r=1$$ 
with
$$
f_0(x,\ve) = \frac{x}{T}, \quad f_1(x,\ve)=0, \quad  F(x, \ve) = 0
$$
in \eqref{yuima:func1}.
So, the call option price requires the composition  of a smooth functional
$$F^\ve(X_t^{(\ve)}) = \frac{1}{T}\int_0^T X_t^{(\ve)} \de t, \qquad r=1$$
with the irregular function
$$
\max(x-K,0)
$$
Monte Carlo methods require a huge number of simulations to get the desired accuracy of the
calculation of the price, while asymptotic expansion of $F^\ve$ provides very accurate approximations.
The \pkg{yuima} package provides functions to construct the functional $F^\ve$, and automatic asymptotic expansion based on Malliavin calculus \citep{Yoshida92b} starting from a \pkg{yuima} object.
Here is an example. Consider a simple geometric Brownian motion of equation \eqref{eqyuima:gbm1} with $\mu=1$ and $X_0=1$. We define the model and the functional below:
<<echo=TRUE, print=FALSE, results=hide>>=
model <- setModel(drift = "x", diffusion = matrix( "x*e", 1,1))
T <- 1
xinit <- 1
K <- 1
f <- list( expression(x/T), expression(0))
F <- 0
e <- 0.5
yuima <- setYuima(model = model, sampling = setSampling(Terminal=T, n=1000))
yuima <- setFunctional( yuima, f=f,F=F, xinit=xinit,e=e)
@
this time the \code{setFunctional} command fills the appropriate slots inside the \code{yuima} object
<<echo=TRUE, print=FALSE>>=
str(yuima@functional)
@
Then, to obtain the first term in the asymptotic expansion (i.e., for $\ve=0$), it is as easy as calling the function \code{F0} on the 
<<echo=TRUE, print=FALSE>>=
F0 <- F0(yuima)
F0
@
so the option price approximation is
<<>>=
max(F0-K,0)  # asian call option price
@
We can go up to the first order approximation adding one term to the expansion
<<echo=TRUE, print=FALSE,results=hide>>=
rho <- expression(0)
epsilon <- e  # noise level
g <- function(x) {
  tmp <- (F0 - K) + (epsilon * x) 
 tmp[(epsilon * x) < (K-F0)] <- 0
 tmp
}
asymp <- asymptotic_term(yuima, block=10, rho, g)
@
and the final value is
<<echo=TRUE, print=FALSE>>=
asymp$d0 + e * asymp$d1  # asymp. exp. of asian call price
@

@@

\subsection{Quasi Maximum Likelihood Estimation}
Consider the multidimensional diffusion process
$$
\de X_t = b(\theta_2,X_t)\de t + \sigma(\theta_1, X_t) \de W_t
$$
where $W_t$ is an {$r$}-dimensional standard Wiener process 
independent of the initial value $X_0=x_0$. 
Quasi-MLE assumes the following approximation of the true log-likelihood for multidimensional diffusions
{\small
\begin{eqnarray}\label{qlik}
\ell_n({\bf X}_n,\theta)
=-\frac12\sum_{i=1}^n\left\{\log\text{det}(\Sigma_{i-1}(\theta_1))
+\frac{1}{\Delta_n}
\Sigma_{i-1}^{-1}(\theta_1)[\Delta X_i-\Delta_n b_{i-1}(\theta_2)]^{\otimes 2}\right\}
\end{eqnarray}}
where $\theta=(\theta_1, \theta_2)$, $\Delta X_i=X_{t_i}-X_{t_{i-1}}$, $\Sigma_i(\theta_1)=\Sigma(\theta_1,X_{t_i})$, $b_i(\theta_2)=b(\theta_2,X_{t_i})$, $\Sigma=\sigma^{\otimes 2}$, \textcolor{red}{$A^{\otimes 2}= A A^T$} and $A^{-1}$ the inverse of $A$, $A[B]^{\otimes 2} = B^T A B$.  Then, \citep[see e.g.,][]{Yoshida92, Kessler97}, the QML estimator of $\theta$ is
\textcolor{red}{$$\hat\theta_n=\arg\max_\theta \ell_n({\bf X}_n,\theta)$$}

@@

@@

\subsection{Adaptive Bayes Estimation}
Consider again the  diffusion process solution to
\begin{equation} 
\de X_t=b(X_t,\theta_2)\de t+\sigma(X_t,\theta_1)\de W_t,
\end{equation}
and the quasi likelihood defined in \eqref{qlik}.

@@

Adaptive Bayes estimation is developed in \pkg{yuima} via the method $\tt adaBayes$.
Consider again the model \eqref{eq:model1} with the same values for the parameters.
In order to perform Bayesian estimation, we need to prepare the prior densities for the parameters.
For simplicity we use uniform distributions in $[0,1]$
<<echo=TRUE>>=
prior <- list(theta2=list(measure.type="code",df="dunif(theta2,0,1)"), theta1=list(measure.type="code",df="dunif(theta1,0,1)"))
@
Then we call $\tt adaBayes$, on the same sample data we used for the \code{qmle} function, as follows
<<echo=TRUE, results=hide>>=
bayes1 <- adaBayes(yuima, start=param.init, prior=prior)
@
and we can compare the adaptive Bayes estimates  with the QMLE estimates
<<echo=TRUE>>=
coef(bayes1)
coef(mle1)
@
%The argument \code{method="nomcmc"} in \code{adaBayes} performs numerical
%integration, otherwise MCMC method is used.
\subsubsection{Small Sample Size}
It is known from the theory that the estimation of the drift in a diffusion process strongly depend on the length of the observation interval $[0,T]$.
In our example above, we took $T=n^\frac13$, with $n = \Sexpr{n}$, which is approximatively  \Sexpr{round(n^(1/3),2)}. Now we reduce the sample size to $n=500$ and the value of $T$ is then $T=\Sexpr{round(500^(1/3),2)}$.
We then apply both quasi-maximum likelihood and adaptive Bayes type estimators to these data
<<>>=
n <- 500
ysamp <- setSampling(Terminal = n^(1/3), n = n)
yuima <- setYuima(model = ymodel, sampling = ysamp)
set.seed(123)
yuima <- simulate(yuima, xinit = 1, true.parameter = list(theta1 = 0.2, theta2 = 0.3))
param.init <- list(theta2=0.5,theta1=0.5)
mle2 <- qmle(yuima, start =param.init , 
lower = list(theta1=0, theta2=0), 
upper = list(theta1=1, theta2=1))
bayes2 <- adaBayes(yuima, start=param.init, prior=prior)
@
and we look at the estimates
<<>>=
coef(bayes2)
coef(mle2)
@
Compared to the results above, we see that the parameters in the diffusion coefficients are estimated with good quality while for the parameters in the drift the quality of estimation deteriorates. The adaptive Bayes estimator seems to perform a little better though.

@@

Let $T \in (0,\infty)$ be a terminal time for possible observations. 
We consider a two dimensional It\^o process $(X^1,X^2)$  
satisfying the stochastic differential equations
\begin{eqnarray*}
\mbox{d}X_t^l &=&\mu_t^l \mbox{d}t + \sigma_t^l \mbox{d}W_t^l ,\quad
t\in[0,T]\\
X_0^l &=& x_0^l        \notag
\end{eqnarray*}
for $l=1,2$. 
Here $W^l$ denote standard Wiener processes with a 
progressively measurable correlation process 
$\mbox{d}\langle W_1 , W_2 \rangle_t = \rho_t \mbox{dt}$,  
$\mu^l_t$ and $\sigma_t^l$ are progressively measurable processes, 
and $x^l_0$ are initial random variables independent of $(W^1,W^2)$.  
Diffusion type processes are in the scope but this model 
can express more sophisticated stochastic structures.

The process $X^l$ is supposed to be observed at 
over the increasing sequence of times 
$T^{l,i}$ $(i\in\mathbb{Z}_{\geq0})$ starting at $0$, up to time T. 
Thus, the observables are $(T^{l,i},X^{l,i})$ with $T^{l,i}\leq T$. 
Each $T^{l,j}$ may be a stopping time, so possibly depends on 
the history of $(X^1,X^2)$ as well as the precedent stopping times.  
Two sequences of stopping times 
$T^{1,i}$ and $T^{2,j}$ are \textit{nonsynchronous},  and 
irregularly spaced, in general. 
In particular, \textcolor{red}{\code{cce}} can apply to estimation of the quadratic variation of a single 
stochastic process sampled regularly/irregularly.

The parameter of interest is the quadratic covariation between $X^1$ and $X^2$:

>>>>

\usepackage{comment}
\excludecomment{en-text}

\usepackage{enumerate}%\begin{enumerate}[{(}i{)}]
\usepackage{bm}
\usepackage{bbm} %\mathbbm{ABCabc\nabla}
\usepackage[bbgreekl]{mathbbol} %\mathbb{abcABC\nabla\bbalpha\bbbeta\Delta  }
\usepackage{natbib}
\usepackage{dsfont}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{latexsym}
\usepackage{amsmath}

@@

From the above, it is possible to see that the jump coefficient is void and the Hurst parameter is set to 0.5, \textcolor{blue}{because this is a model where the driving process is the standard Brownian motion, i.e. a fractional Brownian motion if Hurst index $H=\frac12$.}
Now, with \code{mod1} in hands, it is very easy to simulate a trajectory by  \textcolor{blue}{Euler-Maruyama scheme} of the process as follows
<<echo=TRUE, print=FALSE,fig=TRUE,width=9,height=4,results=hide>>=
set.seed(123)
X <- simulate(mod1)
plot(X)
@

@@

\subsection{Multidimensional Processes}\label{sec:multi}
\textcolor{blue}{Next is an example of a system of two stochastic differential equations for the couple $(X_{t,1}, X_{t,2})$ driven by three independent Brownian motions $(W_{t,1}, W_{t,2}, W_{t,3})$}
$$
\begin{aligned}
\de X_{1,t} &= -3 X_{1,t} \de t + \de W_{1,t} + X_{2,t} \de W_{3,t}\\
\de X_{2,t} &= -(X_{1,t} + 2 X_{2,t}) \de t + X_{1,t} \de W_{1,t} + 3 \de W_{2,t}
\end{aligned}
$$
but this has to be organized into matrix form \textcolor{red}{with a vector of drift expressions and a diffusion matrix}
$$
\left(\begin{array}{c}\de X_{1,t} \\\de X_{2,t}\end{array}\right)=
\left(\begin{array}{c} -3 X_{1,t} \\  -X_{1,t} - 2X_{2,t}\end{array}\right)\de t +
\left(\begin{array}{ccc}1 & 0 & X_{2,t} \\X_{1,t} & 3 & 0\end{array}\right)
\left(\begin{array}{c}\de W_{t,1} \\ \de W_{t,2} \\ \de W_{t,3}\end{array}\right)
$$
\textcolor{blue}{For this system it now necessary to instruct \pkg{yuima} about the state variable on bot the left-hand side of the equation and the right-hand side of the equation, i.e. the \code{solve.variable}.}
<<echo=TRUE, print=FALSE,results=hide>>=
sol <- c("x1","x2") # variable for numerical solution
a <- c("-3*x1","-x1-2*x2")   # drift vector 
b <- matrix(c("1","x1","0","3","x2","0"),2,3)  #  diffusion matrix
mod3 <- setModel(drift = a, diffusion = b, solve.variable = sol)
@
\textcolor{blue}{Looking at the structure of the \code{noise.number} slot in \code{mod3}, one can see that this is now set to 3 which is taken as the number of columns of the diffusion matrix.}
Again, this model can be easily simulated
<<echo=TRUE, print=FALSE,fig=TRUE,width=9,height=4,results=hide>>=
set.seed(123)
X <- simulate(mod3)
plot(X, plot.type="single",lty=1:2)
@

@@

\subsection{Fractional Gaussian Noise}\label{sec:fgn}
The \pkg{yuima} allows for the description of stochastic differential equations driven by fractional Brownian motion of the following type
$$\de X_t =  a(X_t) \de t + b(X_t) \de W_t^H
$$
where $W^H=\left( W^H_t, t\geq 0 \right)$ is a normalized fractional Brownian motion (fBM), {\it i.e.} the zero mean Gaussian processes with covariance function
$$\dE( W_s^H W_t^H )= \frac{1}{2} \left[ |s|^{2H}+|t|^{2H}- |t-s|^{2H}\right]$$ with Hurst exponent $H\in(0,1)$.  The fractional Brownian motion process is  neither Markovian nor a semimartingale for $H\neq\frac{1}{2}$ but
remains Gaussian.  For $H > \frac12$, the solution $X_t$ above  presents the long-range dependance property that makes it useful for different applications in biology, physics, ethernet traffic or finance.
In order to specify a stochastic differential equation driven by fractional Gaussian noise it is necessary to specify the value of the Hurst parameter. For example, if we want to specify the following model
$$\de Y_t =  3 Y_t \de t + \de W_t^H$$
we proceed as follows
<<echo=TRUE, print=FALSE,fig=TRUE,width=9,height=6,results=hide>>=
mod4A <- setModel(drift="3*y", diffusion=1, hurst=0.3, solve.var="y")
mod4B <- setModel(drift="3*y", diffusion=1, hurst=0.7, solve.var="y")
set.seed(123)
X1 <- simulate(mod4A,sampling=setSampling(n=1000))
X2 <- simulate(mod4B,sampling=setSampling(n=1000))
par(mfrow=c(2,1))
par(mar=c(2,3,1,1))
plot(X1,main="H=0.3")
plot(X2,main="H=0.7")
@

@@

\section{Asymptotic Expansion}\label{sec4}
{\colorb For numerical computation of the expectation of a random variable,
the Monte-Carlo method gives a universal solution although it is time consuming and involves 
 stochastic errors of certain scale 
depending on the number of repetitions. 
An alternative method is the asymptotic expansion, that can often give a solution with accuracy comparable or superior to  
Monte-Carlo approaches, 
besides, it has advantage in the computation time because of the approximation by an analytic formula. 
 }

%The \pkg{yuima} 
%package {\colorb provides numerical approximation to the expectation of 
%a functional of a diffusion process in terms of the asymptotic expansion scheme. }
Let us consider a family of 
 $d$-dimensional diffusion processes {\colorb $X=(X^{(\ve)}_t)_{t\in[0,T]}$ ($\ve\in(0,1]$) specified 
 by the stochastic integral equation}
\begin{equation}\label{asy01}
X_t^{(\ve)} = x_0+\int_0^t a(X_s^{(\ve)},\ve)\de s + \int_0^t b(X_s^{(\ve)},\ve)\de W_s, \qquad t\in[0,T]
\end{equation}
{\colorb for $\ve \in(0,1]$, 
where}  $W_t=(W_{1,t}, \ldots, W_{r,t})$ \textcolor{red}{is an} $r$-dimensional Wiener process.
The functional of interest is expressed in the following abstract form
\begin{equation}
\label{yuima:func1}
F^{(\ve)}  = \sum_{\alpha=0}^r \int_0^T f_\alpha(X_t^{(\ve)},\ve)\de W_t^\alpha + F(X_T^{(\ve)},\ve), \qquad W_t^0=t.
\end{equation}

A typical application is the Asian option pricing. For example, in the Black-Scholes model
\begin{equation}
\label{eqyuima:gbm1}
\de X_t^{(\ve)} = \mu X_t^{(\ve)}\de t + \ve X_t^{(\ve)} \de W_t, 
\end{equation}
the price of the option under zero interest rate is of the form
$$ \dE\left[\max\left( \frac{1}{T}\int_0^T X_t^{(\ve)} \de t - K,0\right)\right].$$
%
Thus the functional of interest is
$$F^{(\ve)} = \frac{1}{T}\int_0^T X_t^{(\ve)} \de t, \qquad r=1$$ 
with
$$
f_0(x,\ve) = \frac{x}{T}, \quad f_1(x,\ve)=0, \quad  F(x, \ve) = 0
$$
in \eqref{yuima:func1}.
%
%\begin{en-text}%%%%%%%%%%%
%So, the call option price requires the composition  of a smooth functional
%$$F^\ve(X_t^{(\ve)}) = \frac{1}{T}\int_0^T X_t^{(\ve)} \de t, \qquad r=1$$
%with the irregular function
%$$
%\max(x-K,0)
%$$
%\end{en-text}%%%%%%%%%%
%
{\colorb Similarly, for $F(x,\ve)=x$, 
the functional becomes $F^{(\ve)}=X^{(\ve)}_T$ and the price of the European call option is 
$\dE[\max(X^{(\ve)}_T-K,0)]$. This value has a closed form in the Black-Sholes economy but 
it is necessary to apply some numerical method for pricing the Asian option even in this linear case. }

{\colorb 
Returning to the general system (\ref{asy01})-(\ref{yuima:func1}), 
we will assume that the stochastic system is deterministic in the limit as $\ep\downarrow0$, 
that is,  
$$ 
b(\cdot,0)\>=\>0 \quad \mbox{and}\quad  f_\alpha(\cdot,0)=0\quad (\alpha=1,...,r).
$$ 
%
Since $X^{(0)}_t$ is deterministic solution to the ordinary differential equation 
$$
\de X^{(0)}_t/\de t =a(X^{(0)}_t,0),\quad X^{(0)}_0=x_0,
$$
the functional $F^{(0)}$ becomes a constant given by 
\begin{equation} \label{130106-1}
F^{(0)} = \int_0^T f_0(X^{(0)}_t,0)\de t+F(X^{(0)}_T,0).
\end{equation}

Under standard regularity of $a$, $b$, $f_\alpha$ and $F$, 
it is possible to show $F^{(\ep)}$ has a version that is smooth in $\ep\in[0,1)$ almost surely, and hence, 
$$
\tilde{F}^{(\ep)} := \ep^{-1}(F^{(\ep)}-F^{(0)})
$$
admits a stochastic expansion
\begin{equation}
\label{130206-8}
\tilde{F}^{(\ep)} \quad \sim \quad
\tilde{F}^{[0]}+\ep\tilde{F}^{[1]}+\ep^2\tilde{F}^{[2]}+\cdots
\end{equation}
as $\ep\downarrow0$. 
This stochastic expansion makes sense in the Sobolev spaces of the Malliavin calculus. 
Then the so-called Watanabe's theory (\cite{Watanabe87}) validates 
the asymptotic expansion of the (generalized) expectation 
\begin{equation} \label{130206-2}
\dE[g(\tilde{F}^{(\ep)})] \quad \sim \quad d_0(g)
+\ep d_1(g)+\ep^2 d_2(g)+\cdots
\end{equation} 
as $\ep\downarrow0$ for measurable functions $g$ at most polynomial growth or 
more generally for Schwartz distributions, 
under the uniform nondegeneracy of the Malliavain covariance of $\tilde{F}^{(\ep)}$.
\footnote{This condition ensures smoothness of the distribution of $\tilde{F}^{(\ep)}$. 
It should be remarked that the Watanabe's theory is more general than the present use 
for the variable $\tilde{F}^{(\ep)}$ having a Gaussian principal part $\tilde{F}^{(0)}$. }
%
In the present situation, each $d_i(g)$ is expresses as 
$$
d_i(g) = \int g(z)p_i(z)\phi(z;0,v)dz,
$$
where $p_i$ is a polynomial and $\phi(z;0,v)$ is the density of the normal distribution $N(0,v)$
with $v=\mbox{Cov}[\tilde{F}^{(0)}]$. 
Polynomials $p_i$ are given by the conditional expectation of multiple Wiener integrals. 
%This expansion formula is implemented in \pkg{yuima}. 
%In finance, we can apply this expansion to $g(z)=\max(z-c,0)$, for example. 
The expansion (\ref{130206-2}) holds uniformly in a class of functions $g$.

As mentioned above, 
Monte Carlo methods require a huge number of simulations to get the desired accuracy of the
calculation of the expectation, while the asymptotic expansion of $F^{(\ve)}$ gives very fast 
and accurate approximation by analytic formulation.
The \pkg{yuima}  package provides functions to construct the functional $F^{(\ve)}$, and automatic asymptotic expansion based on the Malliavin calculus.  starting from a  \pkg{yuima} object. 
This asymptotic expansion approach to option pricing was proposed in \citep{Yoshida92b} 
and many related papers are available today.

Though our method can be applied to the nonlinear system (\ref{asy01})-(\ref{yuima:func1}), 
just for an example, we shall consider 
the Asian call option of the geometric Brownian motion of equation \eqref{eqyuima:gbm1} with $\mu=1$ and $x_0=1$, 
and 
\begin{equation} \label{130206-5} 
g(x)= %F^{(0)}-K+\ep x\>=\>\ep\left(
\max\left(F^{(0)}-K+\ep x,0\right)%\right). 
\end{equation} 
Set the model  (\ref{asy01}) and the functional (\ref{yuima:func1}) as follows:
<<echo=TRUE, print=FALSE, results=hide>>=
model <- setModel(drift = "x", diffusion = matrix( "x*e", 1,1))
T <- 1
xinit <- 1
K <- 1
f <- list( expression(x/T), expression(0))
F <- 0
e <- 0.5
yuima <- setYuima(model = model, sampling = setSampling(Terminal=T, n=1000))
yuima <- setFunctional( yuima, f=f,F=F, xinit=xinit,e=e)
@
This time the \code{setFunctional} command fills the appropriate slots inside the \code{yuima} object
<<echo=TRUE, print=FALSE>>=
str(yuima@functional)
@
Then the limit $F^{(0)}$ of $F^{(\ep)}$ is easily obtained by calling the function \code{F0} on the \code{yuima} object:
<<echo=TRUE, print=FALSE>>=
F0 <- F0(yuima)
F0
@
Set the function $g$ according to (\ref{130206-5}): 
%<<>>=
%max(F0-K,0)  # asian call option price
%@
%We can go up to the first order approximation adding one term to the expansion
<<echo=TRUE, print=FALSE,results=hide>>=
rho <- expression(0)
epsilon <- e  # noise level
g <- function(x) {
  tmp <- (F0 - K) + (epsilon * x) 
 tmp[(epsilon * x) < (K-F0)] <- 0
 tmp
}
@
Now we are at the point of computing the coefficients $d_i$ ($i=0,1,2$) in the expansion of 
the price $\dE[\max(F^{(\ep)}-K,0)]$ 
by applying the function \code{asymptotic_term}: 
<<echo=TRUE, print=FALSE,results=hide>>=
asymp <- asymptotic_term(yuima, block=10, rho, g)
asymp
@
Then the sums 
<<echo=TRUE, print=FALSE>>=
asymp$d0 + e * asymp$d1  # 1st order asymp. exp. of asian call price
asymp$d0 + e * asymp$d1+ e^2* asymp$d2  # 2nd order asymp. exp. of asian call price
@
give the first and second order asymptotic expansions, respectively.

{\coloro koko MC here!! Why don't we make asyexp = \code{asymptotic_term} + sums? 
Also Monte-Carlo driver montecarlo = \code{simulate} with number of repetition 
+ plot + comparison of plot and theoretical curve?}

We remark that the expansion of $\dE[g(\tilde{F}^{(\ep)})G^{(\ep)}]$ is also possible by the same method 
for a functional $G^{(\ep)}$ having a stochastic expansion like (\ref{130206-8}). 
Thus our method works even under the existence of a stochastic discount factor.

@@

\subsection{Quasi Maximum Likelihood Estimation}
{\bf[please remember: $\hat\theta$ = MLE;  $\tilde \theta$ = BAYES; $\check\theta$ = LASSO]}
\par
Consider the multidimensional diffusion process
$$
\de X_t = a(X_t,\theta_2)\de t + b(X_t,\theta_1) \de W_t
$$
where $W_t$ is an {$r$}-dimensional standard Wiener process 
independent of the initial value $X_0=x_0$. 
Quasi-MLE assumes the following approximation of the true log-likelihood for multidimensional diffusions
{\small
\begin{eqnarray}\label{qlik}
\ell_n({\bf X}_n,\theta)
=-\frac12\sum_{i=1}^n\left\{\log\text{det}(\Sigma_{i-1}(\theta_1))
+\frac{1}{\Delta_n}
\Sigma_{i-1}^{-1}(\theta_1)[\Delta X_i-\Delta_n a_{i-1}(\theta_2)]^{\otimes 2}\right\}
\end{eqnarray}}
where $\theta=(\theta_1, \theta_2)$, $\Delta X_i=X_{t_i}-X_{t_{i-1}}$, $\Sigma_i(\theta_1)=\Sigma(\theta_1,X_{t_i})$, $a_i(\theta_2)=a(X_{t_i},\theta_2)$, $\Sigma=b^{\otimes 2}$, \textcolor{red}{$A^{\otimes 2}= A A^T$} and $A^{-1}$ the inverse of $A$, $A[B]^{\otimes 2} = B^T A B$.  Then, \citep[see e.g.,][]{Yoshida92, Kessler97}, the QML estimator of $\theta$ is
\textcolor{red}{$$\hat\theta_n=\arg\max_\theta \ell_n({\bf X}_n,\theta)$$}

@@

\subsubsection{QMLE in practice}

@@

\subsection{Adaptive Bayes Estimation}
Consider again the  diffusion process solution to
\begin{equation} 
\de X_t=a(X_t,\theta_2)\de t+b(X_t,\theta_1)\de W_t,
\end{equation}
and the quasi likelihood defined in \eqref{qlik}.

@@

Adaptive Bayes estimation is developed in \pkg{yuima} via the method $\tt adaBayes$.
Consider again the model \eqref{eq:model1} with the same values for the parameters.
In order to perform Bayesian estimation, we need to prepare the prior densities for the parameters.
For simplicity we use uniform distributions in $[0,1]$
<<echo=TRUE>>=
prior <- list(theta2=list(measure.type="code",df="dunif(theta2,0,1)"), theta1=list(measure.type="code",df="dunif(theta1,0,1)"))
@
Then we call $\tt adaBayes$, on the same sample data we used for the \code{qmle} function, as follows
<<echo=TRUE, results=hide>>=
bayes1 <- adaBayes(yuima, start=param.init, prior=prior)
@
and we can compare the adaptive Bayes estimates  with the QMLE estimates
<<echo=TRUE>>=
coef(bayes1)
coef(mle1)
@
%The argument \code{method="nomcmc"} in \code{adaBayes} performs numerical
%integration, otherwise MCMC method is used.
\subsubsection{The Effect of Small Sample Size in Drift Estimation}
It is known from the theory that the estimation of the drift in a diffusion process strongly depend on the length of the observation interval $[0,T]$.
In our example above, we took $T=n^\frac13$, with $n = \Sexpr{n}$, which is approximatively  \Sexpr{round(n^(1/3),2)}. Now we reduce the sample size to $n=500$ and the value of $T$ is then $T=\Sexpr{round(500^(1/3),2)}$.
We then apply both quasi-maximum likelihood and adaptive Bayes type estimators to these data
<<>>=
n <- 500
ysamp <- setSampling(Terminal = n^(1/3), n = n)
yuima <- setYuima(model = ymodel, sampling = ysamp)
set.seed(123)
yuima <- simulate(yuima, xinit = 1, true.parameter = list(theta1 = 0.2, theta2 = 0.3))
param.init <- list(theta2=0.5,theta1=0.5)
mle2 <- qmle(yuima, start =param.init , 
lower = list(theta1=0, theta2=0), 
upper = list(theta1=1, theta2=1))
bayes2 <- adaBayes(yuima, start=param.init, prior=prior)
@
and we look at the estimates
<<>>=
coef(bayes2)
coef(mle2)
@
Compared to the results above, we see that the parameters in the diffusion coefficients are estimated with good quality while for the parameters in the drift the quality of estimation deteriorates. The adaptive Bayes estimator seems to perform a little better though.

@@

Let $T \in (0,\infty)$ be a terminal time for possible observations. 
We consider a two dimensional It\^o process $(X_1,X_2)$  
satisfying the stochastic differential equations
\begin{eqnarray*}
\mbox{d}X_{l,t} &=&\mu_{l,t} \mbox{d}t + \sigma_{l,t} \mbox{d}W_{l,t} ,\quad
t\in[0,T]\\
X_{l,0} &=& x_{l,0}        \notag
\end{eqnarray*}
for $l=1,2$. 
Here $W_l$ denote standard Wiener processes with a 
progressively measurable correlation process 
$\mbox{d}\langle W_1 , W_2 \rangle_t = \rho_t \mbox{dt}$,  
$\mu_{l,t}$ and $\sigma_{l,t}$ are progressively measurable processes, 
and $x_{l,0}$ are initial random variables independent of $(W_1,W_2)$.  
Diffusion type processes are in the scope but this model 
can express more sophisticated stochastic structures.

The process $X_l$ is supposed to be observed at 
over the increasing sequence of times 
$T^{l,i}$ $(i\in\mathbb{Z}_{\geq0})$ starting at $0$, up to time T. 
Thus, the observables are $(T^{l,i},X_{l,i})$ with $T^{l,i}\leq T$. 
Each $T^{l,j}$ may be a stopping time, so possibly depends on 
the history of $(X_1,X_2)$ as well as the precedent stopping times.  
Two sequences of stopping times 
$T^{1,i}$ and $T^{2,j}$ are \textit{nonsynchronous},  and 
irregularly spaced, in general. 
In particular, \textcolor{red}{\code{cce}} can apply to estimation of the quadratic variation of a single 
stochastic process sampled regularly/irregularly.

The parameter of interest is the quadratic covariation between $X_1$ and $X_2$:

\begin{equation}
\theta=
\langle X_1 , X_2 \rangle_T = \int_0^T \sigma_{1,t} \sigma_{2,t} \rho_t \mbox{d}t.
\end{equation}
The target variable $\theta$ is random in general.
